The role of priors, competence and relevance for the interpretation of disjunction \thanks{We thank \ldots}

### 2022-01-16 analysis/main/03_xor-some_prereg_analysis_final.Rmd
address MFs comments, update plots, grid search

@@

ggsave("../../writing/ELM2/posterior-effects.pdf", width=7, height=5)
```

@@

# check P that the effects are positive / negative / no effect present
posterior_hypotheses_cat <- model_xor_cat_zScored_posteriors %>% 
  select(prior_xor, prior_some, 
         competence_xor, competence_some,
         relevance_xor, relevance_some) %>%
  gather(key, val) %>%
  group_by(key) %>% mutate(positive = mean(val > 0.05),
                           negative = mean(val < -0.05),
                           no = mean(val %>% between(-0.05, 0.05))) %>%
  summarise(positive_eff = mean(positive),
            negative_eff = mean(negative),
            no_eff = mean(no))
posterior_hypotheses_cat
```
```{r}
test_conjunction_of_all_hypotheses(posterior_hypotheses_cat)
```

@@

model_xor_cat_posteriors_long %>%
  ggplot(.) + # 
  stat_halfeye(aes(y = effect, x = value, 
                   fill = stat(case_when(x %>% between(-0.05, 0.05) ~ "no effect", x < -0.05 ~ "negative effect", x > 0.05 ~ "positive effect"))
                   ), 
               alpha = .7) +
  geom_vline(xintercept = c(-.05, .05), linetype = "dashed") +
  scale_fill_manual(name = "Direction of effect", values = c("coral", "yellow3", "mediumseagreen")) +
  ggtitle("Posterior effect distributions. Points indicate posterior means\nThick lines indicate 66% CrIs, thin lines 95% CrIs.")

@@

model_mean_posteriors_long %>%
  ggplot(.) + # 
  stat_halfeye(aes(y = effect, x = value, 
                   fill = stat(case_when(x %>% between(-0.05, 0.05) ~ "no effect", x < -0.05 ~ "negative effect", x > 0.05 ~ "positive effect"))
                   ), 
               alpha = .7) +
  geom_vline(xintercept = c(-.05, .05), linetype = "dashed") +
  scale_fill_manual(name = "Direction of effect", values = c("coral", "yellow3", "mediumseagreen")) +
  ggtitle("Posterior effect distributions. Points indicate posterior means\nThick lines indicate 66% CrIs, thin lines 95% CrIs.")

>>>>

For the paper: Investigate whether we get the same results (qualitatively) if we set the ROPE to 0:
```{r}
posterior_hypotheses_nullROPE <- model_SI_posteriors %>% 
  select(prior_xor, prior_some, 
         competence_xor, competence_some,
         relevance_xor, relevance_some) %>%
  gather(key, val) %>%
  group_by(key) %>% mutate(positive = mean(val > 0),
                           negative = mean(val < 0)) %>% # no = mean(val = 0)
  summarise(positive_eff = mean(positive),
            negative_eff = mean(negative) # no_eff = mean(no)
            )
posterior_hypotheses_nullROPE
```

@@

# ggsave("../../writing/ELM2/posterior-effects.pdf", width=7, height=5)
```

@@

# check P that the effects are positive / negative / no effect present
posterior_hypotheses_cat <- model_xor_cat_zScored_posteriors %>% 
  select(prior_xor, prior_some, 
         competence_xor, competence_some,
         relevance_xor, relevance_some) %>%
  gather(key, val) %>%
  group_by(key) %>% mutate(positive = mean(val > 0.05),
                           negative = mean(val < -0.05),
                           no = mean(val %>% between(-0.05, 0.05))) %>%
  summarise(positive_eff = mean(positive),
            negative_eff = mean(negative),
            no_eff = mean(no))
posterior_hypotheses_cat
```

```{r}
test_conjunction_of_all_hypotheses(posterior_hypotheses_cat)
```

@@

model_xor_cat_posteriors_long %>%
  mutate(effect = factor(effect, 
                         levels = c("relevance_xor", "relevance_some", "prior_xor", "prior_some", "competence_xor", "competence_some"), 
                         labels = c("Relevance (or)", "Relevance (some)", "Prior (or)", "Prior (some)", "Competence (or)", "Competence (some)"))) %>%
  ggplot(.) + # 
  stat_halfeye(aes(y = effect, x = value, 
                   fill = stat(case_when(x %>% between(-0.05, 0.05) ~ "no effect", x < -0.05 ~ "negative effect", x > 0.05 ~ "positive effect"))
                   ), 
               alpha = .7) +
  geom_vline(xintercept = c(-.05, .05), linetype = "dashed") +
  scale_fill_manual(name = "Direction of effect", values = c("coral", "yellow3", "mediumseagreen")) +
  labs(x = "Effect size", y = "Effect") +
  theme(legend.position = "top",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        strip.text.x = element_text(size = 13),
        strip.text.y = element_text(size = 13))

#ggsave("../../writing/paper/images/posterior-effects-categorical.pdf", width=7, height=5)
```

@@

model_mean_posteriors_long %>%
  mutate(effect = factor(effect, 
                         levels = c("relevance_xor", "relevance_some", "prior_xor", "prior_some", "competence_xor", "competence_some"), 
                         labels = c("Relevance (or)", "Relevance (some)", "Prior (or)", "Prior (some)", "Competence (or)", "Competence (some)"))) %>%
  ggplot(.) + # 
  stat_halfeye(aes(y = effect, x = value, 
                   fill = stat(case_when(x %>% between(-0.05, 0.05) ~ "no effect", x < -0.05 ~ "negative effect", x > 0.05 ~ "positive effect"))
                   ), 
               alpha = .7) +
  geom_vline(xintercept = c(-.05, .05), linetype = "dashed") +
  scale_fill_manual(name = "Direction of effect", values = c("coral", "yellow3", "mediumseagreen")) +
  labs(x = "Effect size", y = "Effect") +
  theme(legend.position = "top",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        strip.text.x = element_text(size = 13),
        strip.text.y = element_text(size = 13))

### 2022-01-16 writing/paper/paper_draft.tex
address MFs comments, update plots, grid search

More specifically, we conducted a simulation-based Bayesian power analysis which aimed at determining how many participants are required in order to detect a theoretically motivated conjunctive effect of the predictors prior, relevance and competence with a confidence of at least 0.8 \citep[e.g.,][]{degen2015investigating}. The simulations were based on an assumed effect size $\beta = 0.15$ for all predictors, and a Region Of Practical Equivalence (ROPE) $\delta = 0.05$ for judging evidence for the directionality of a coefficient given the data \citep{kruschke2014doing}. The simulations were further based on the maximal desired model justified theoretically and by the design.\footnote{The maximal model in R syntax is: \texttt{rating = trigger * prior * competence * relevance}. The random effects were omitted for computational tractability reasons and due to the lack of a well-founded estimates of group-level effect sizes.}

@@

We focus on six slope coefficients of the regression model, for the effects of relevance, competence and prior, respectively, once for ``some'' and once for ``or''. We check whether the posterior estimate of each effect is in either one of three intervals: (1) negative effect (<= -0.05), (2) no effect (between -0.05 and 0.05), or (3) positive effect (>= 0.05). 
We interpret the data as providing evidence in favor of an effect (positive, negative, no effect) if the posterior probability of the effect being true is >= 0.95 (i.e., 95\% of posterior samples are in the corresponding interval). In particular, we speak of evidence in favor of the scalar implicature account of the exclusive disjunction interpretation if the posterior probability of the prior effect being negative, the relevance and competence effects being positive is >= 0.95.

\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-main.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Main analysis results: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). \mf{better use nicer labels for $y$-axis: use ``or'' not ``xor'' and no hyphens?; same for next plots}}
	\label{posteriors-main}
\end{figure*}

Consistent with predictions, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true, as inidcated by the probability of the negative effect of prior being $P = 0.999$ (Fig.\ref{posteriors-main}, \texttt{prior\_some} line, orange color). Similarly, we find a clear positive effect for speaker competence ($P =  1$, Fig.\ref{posteriors-main}, \texttt{competence\_some} line, green color). However, we do not find any clear effects of relevance (Fig.\ref{posteriors-main}, \texttt{relevance\_some} line, split colors). If anything, the data supports the hypothesis that the effect is not negative.
We hypothesize that the absense of a clear-cut relevance effect might be due to the difficulty manipulating this dimension contextually. It might be that the vignettes used in the experiment failed to make the listener relevance salient enough for any effects to be borne out. Further experiments should improve this manipulation, as well as additionally consider speaker relevance.
\mf{best first just state the results and use the last two sentences for ``Discussion'' section}

For the trigger ``or'', in contrast, we only find a positive effect of competence ($P =  0.993$, Fig.\ref{posteriors-main}, \texttt{competence\_xor} line, green color).  We find no comprehensive effects for relevance or competence. While the reason for the former might be the same as for ``some'', we suggest that these results might also be influenced by collinearity of the effects. We turn to this conjecture in light of the exploratory analyses, described below. 
Overall, these results would provide evidence against the scalar implicature hypothesis because not all factors yielded an effect on the robustness of scalar implicatures generated from ``some'' and ``or''. Similarly, the results would provide evidence against the identity hypothesis, as an effect of competence \mf{prior?} was observed for ``some'', but not for ``or''.\mf{we should explain somewhere where the threshold of 0.05 comes from; when discussing the results here we could add that (think) no effects becomes credible even if the threshold would be reduced to 0}

\subsection{Exploratory analysis}
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-categorical.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with categorical predictors: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-cat}
\end{figure*}
In addition to the main analysis, we conducted two exploratory analyses. First, we fit a model using categorical predictors only. That is, we fit a Bayesian linear mixed effects regression model, regressing the inference rating against our prior intuitive classifications of the vignettes on the dimensions relevance, competence and prior (high or low, respectively), the effect of trigger, and their interactions. The random effects structure and the priors were identical to the main model. The factors were sum-coded, -1 representing the low condition, 1 representing the high condition. 
This model decouples the predictors by dichotomizing their classification and, therefore, addresses the potential collinearty issue. Furthermore, it provides a comparison between participants' judgements and our intuitive prior classifications of the vignettes.

Given this model and the data, we found very similar results for both triggers: We observe a negative effect of prior and a positive effect of competence, in line with predictions of the scalar implicature hypothesis (Fig.\ref{posteriors-cat}). However, we again observe no effects of relevance. If anything, the relevance effect tends in the direction opposite to predictions of the SI account for the trigger ``or''. 
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-means.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with predictor ratings averaged by-vignette: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-mean}
\end{figure*}

@@

\section{Discussion}
In sum, the results provide moderate evidence for the scalar implicature account of ``some'' and ``or''. While we could confirm the effects of prior and comptence for the interpretation of ``some'' as predicted by the SI hypothesis, we could not verify them for the disjunction. We also did not find evidence for effects of relevance on the robustness of scalar implicature generation, given the vignettes. Furthermore, our results hint at intriguing dependencies between relevance and prior effects for the interpretation of exclusive disjunctions. Taken together, the results might speak against the identity hypothesis, indicating that there might be subtle differences in the conceptualization of the stronger alternative for ``some'' and ``or''.

\section{Appendix}
\subsection{Vignettes}
Example transcript of an “or”-story and the associated statements:

Story: “Bill and Eric went to a bar. Eric’s wife, Lily, promised to pick them up if Eric drank any alcohol. She phones up Bill to ask about their night.”
Example comprehension statements: “Eric and Lily went to a bar together.” (false), “Eric and Bill went to see a football game.” (false), “Eric and Bill go out to a bar every weekend.” (uncertain), “Eric's wife has a driver’s licence.” (true)
Competence statement: “Bill knows whether Eric drank both wine and vodka.”
Relevance statement: “It is important for Lily to know whether her husband drank both wine and vodka.”
Prior statements: “If Eric drank wine, it is likely that he drank vodka as well.”, “If Eric drank vodka, it is likely that he drank wine as well.”
Critical utterance: “Bill tells Lily: 'Your husband drank wine or vodka.'”
Implicature statement: “From what Bill said we may conclude that Eric did not drink both wine and vodka.”

Example transcript of a “some”-story and the associated statements:

Story: “Bernard just finished his final exam. He isn't the brightest student, but if he passes this one, he will be through to the next grade. During the break, he goes to the teacher's lounge. He cannot wait to ask his teacher whether he passed. Bernard's teacher only just started correcting his exam.”
Example comprehension statements: “Bernard wants to get to the next grade.” (true), “Bernard wants to know his grade for his college application.” (false), “Bernard took a final exam in biology.” (uncertain), “Bernard is not the best student of his class.” (true)
Competence statement: “Bernard’s teacher knows whether all of the answers Bernard gave on the exam are correct.”
Prior statement: “If at least some of the answers Bernard gave on the exam are correct, then all of them are correct.”
Relevance statement: “It is important for Bernard to know whether all of the answers he gave on the exam are correct.”
Critical utterance: “Bernard’s teacher says: ‘Some of the answers you gave are correct.’”
Implicature statement: “From what Bernard’s teacher said we may conclude that not all of Bernard’s answers are correct.”

>>>>

More specifically, we conducted a simulation-based Bayesian power analysis which aimed at determining how many participants are required in order to detect a theoretically motivated conjunctive effect of the predictors prior, relevance and competence with a confidence of at least 0.8 \citep[e.g.,][]{degen2015investigating}. The simulations were based on an assumed effect size $\beta = 0.15$ for all predictors, and a Region Of Practical Equivalence (ROPE) $\delta = 0.05$ for judging evidence for the directionality of a coefficient given the data \citep{kruschke2014doing}. The simulations were further based on the maximal desired model justified theoretically and by the design.\footnote{The maximal model in R syntax is: \texttt{rating = trigger * prior * competence * relevance}. The random effects were omitted for computational tractability reasons and due to the lack of well-founded estimates of group-level effect sizes.}

@@

We focus on six slope coefficients of the regression model, for the effects of relevance, competence and prior, respectively, once for ``some'' and once for ``or''. We check whether the posterior estimate of each effect is in either one of three intervals: (1) negative effect (<= -0.05), (2) no effect (between -0.05 and 0.05), or (3) positive effect (>= 0.05). We set the threshold for considering an effect as positive or negative to ±0.05 because we consider 0.05 to be the Region of Practical Equivalence (ROPE) for the effect sizes we expect based on literature \citep{kruschke2014doing}.
We interpret the data as providing evidence in favor of an effect (positive, negative, no effect) if the posterior probability of the effect being true is >= 0.95 (i.e., 95\% of posterior samples are in the corresponding interval). The probabilities of the respective coefficients lying in a particular interval are reported below (i.e., for instance, if $P=0.95$ is reported, it means that 95\% of the posterior samples of the given coefficient are in the respective interval). In particular, we speak of evidence in favor of the scalar implicature account of the exclusive disjunction interpretation if the posterior probability of the prior effect being negative, the relevance and competence effects being positive is >= 0.95.

\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-main.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Main analysis results: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals.}
	\label{posteriors-main}
\end{figure*}

Consistent with predictions, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true, as inidcated by the probability of the negative effect of prior being $P = 0.999$ (Fig.\ref{posteriors-main}, \texttt{Prior~(some)} line, orange color). Similarly, we find a clear positive effect for speaker competence ($P =  1$, Fig.\ref{posteriors-main}, \texttt{Competence~(some)} line, green color). However, we do not find any clear effects of relevance (Fig.\ref{posteriors-main}, \texttt{Relevance~(some)} line, split colors). If anything, the data supports the hypothesis that the effect is not negative.

For the trigger ``or'', in contrast, we only find a positive effect of competence ($P =  0.993$, Fig.\ref{posteriors-main}, \texttt{Competence~(or)} line, green color).  We find no comprehensive effects for relevance or competence. While the reason for the former might be the same as for ``some'', we suggest that these results might also be influenced by collinearity of the effects. We turn to this conjecture in light of the exploratory analyses, described below. 
Overall, these results would provide evidence against the scalar implicature hypothesis because not all factors yielded an effect on the robustness of scalar implicatures generated from ``some'' and ``or''. Similarly, the results would provide evidence against the identity hypothesis, as an effect of prior was observed for ``some'', but not for ``or''.

\subsection{Exploratory analysis}
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-categorical.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with categorical predictors: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals. \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-cat}
\end{figure*}
In addition to the main analysis, we conducted two exploratory analyses. First, we fit a model using categorical predictors only. That is, we fit a Bayesian linear mixed effects regression model, regressing the inference rating against our prior intuitive classifications of the vignettes on the dimensions relevance, competence and prior (high or low, respectively), the effect of trigger, and their interactions. The random effects structure and the priors were identical to the main model. The factors were sum-coded, -1 representing the low condition, 1 representing the high condition. 
This model decouples the predictors by dichotomizing their classification and, therefore, addresses the potential collinearty issue. Furthermore, it provides a comparison between participants' judgements and our intuitive prior classifications of the vignettes.

Given this model and the data, we found very similar results for both triggers: We observe a negative effect of prior and a positive effect of competence, in line with predictions of the scalar implicature hypothesis (Fig.\ref{posteriors-cat}). However, we again observe no effects of relevance. If anything, the relevance effect tends in the direction opposite to predictions of the SI account for the trigger ``or''. 
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-means.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with predictor ratings averaged by-vignette: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals. \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-mean}
\end{figure*}

@@

Finally, we conduct an exploratory analysis when setting the ROPE to 0. That is, we investigate whether the posterior estimates for the regression coefficients of interest are positive ($>0$) or negative ($<0$) and interpret the data as providing evidence in favor of a particular direction if at least 95\% of posterior samples for the respective coefficent are positive or negative. As in the main analysis, we find credible positive effects of competence for both ``some'' ($P = 1$) and ``or'' ($P = 1$), and a credible negative effect of prior for ``some'' ($P = 1$), but not for ``or'' ($P=0.91$). However, in contrast to the main analysis, we do find a credible positive effect of relevance for ``or'' ($P=0.978$), but not for ``some'' ($P=0.829$). \pt{We should include justification for using a ROPE in the first place here}

\section{Discussion}
In sum, the results provide moderate evidence for the scalar implicature account of ``some'' and ``or''. While we could confirm the effects of prior and comptence for the interpretation of ``some'' as predicted by the SI hypothesis, we could not verify them for the disjunction. We also did not find evidence for effects of relevance on the robustness of scalar implicature generation, given the vignettes. We hypothesize that the absense of a clear-cut relevance effect might be due to the difficulty manipulating this dimension contextually. It might be that the vignettes used in the experiment failed to make the listener relevance salient enough for any effects to be borne out. Further experiments should improve this manipulation, as well as additionally consider speaker relevance.

Furthermore, our results hint at intriguing dependencies between relevance and prior effects for the interpretation of exclusive disjunctions. Taken together, the results might speak against the identity hypothesis, indicating that there might be subtle differences in the conceptualization of the stronger alternative for ``some'' and ``or''.

\section{Appendix}
\subsection{Vignettes}
\pt{Include all the vignettes?}
